\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}
\usepackage{float}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,calc}
\pgfplotsset{compat=1.18}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue
}

% --- Rust listings (minimal, no external deps) ---
\lstdefinelanguage{Rust}{
  keywords={as, break, const, continue, crate, else, enum, extern, false, fn, for, if, impl, in, let, loop, match, mod, move, mut, pub, ref, return, self, Self, static, struct, super, trait, true, type, unsafe, use, where, while, async, await, dyn},
  sensitive=true,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]",
}

\lstset{
  language=Rust,
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  breaklines=true,
  showstringspaces=false,
  keywordstyle=\bfseries\color{blue!60!black},
  commentstyle=\color{green!40!black},
  stringstyle=\color{red!60!black}
}

\title{Medical Care Robot Coordination System (MCRoCS, ``McRocks''): Written Report}
\author{FA, Haochen\\Student ID: 24113347D\\COMP2432 Operating Systems}
\date{25/26 Sem 2}

\begin{document}
\maketitle

\begin{abstract}
Medical Care Robot Coordination System (MCRoCS, ``McRocks'') is a lightweight operating-system core that coordinates medical-care robots with safe concurrency. It implements exactly the three Project Blaze services: a thread-safe task queue, exclusive zone access control, and a heartbeat-based health monitor. The main challenge was preserving correctness under concurrent threads while keeping the synchronization logic small enough to explain and test. The design uses Rust's \texttt{Mutex} and \texttt{Condvar} for shared-state coordination, while atomics are reserved for simulation metrics and validation \citep{rustmutex,rustcondvar}. The CLI demo deterministically shows concurrent task requests, non-overlapping zone occupancy, and an offline robot detected after missed heartbeats. Fresh release-mode benchmarks report throughput, zone-wait latency, and CPU time across contention, scalability, and offline scenarios; tests verify no duplicate task consumption, zone exclusivity, and timeout detection. The result is a minimal, explainable coordination layer aligned with the project's operating-systems objectives.
\end{abstract}

\section{Introduction}
Modern hospitals increasingly rely on autonomous robots for delivery, sanitation, and assistance. In such settings, coordination errors are not merely inconveniences; they can cause safety incidents, resource contention, or service delays. Medical Care Robot Coordination System addresses this by building a minimal operating-system core that enforces safe concurrency without introducing complex scheduling or preemption. The goal is not to model a complete robot control stack but to implement and demonstrate three essential coordination services under concurrent load: a task queue that ensures each task is assigned at most once, a zone access controller that prevents two robots from occupying the same physical zone simultaneously, and a health monitor that detects robots that stop sending heartbeats.

The project objectives are closely tied to operating-systems fundamentals. First, it must implement correct concurrency control so that shared state remains consistent under multiple threads. Second, it must use explicit synchronization to prevent race conditions. Third, it must coordinate the interactions among threads to produce a predictable, observable outcome that can be demonstrated in a short runtime. The minimal scope is intentional: by removing preemption and complex scheduling policies, the design keeps focus on safety invariants and synchronization correctness. This also makes it possible to reason about system state transitions and to write tests that are both meaningful and deterministic.

The implementation is a Rust Cargo project organized around three core modules, plus a simulation driver. The core coordination path uses \texttt{Mutex} and \texttt{Condvar} with narrow lock scopes; atomics appear only in the simulation harness for metrics and validation. Rust's ownership model and standard library concurrency support help enforce data-race freedom by construction \citep{klabnik2022}. The simulation spawns multiple robot threads, drives task consumption, enforces zone ownership, and checks for offline robots in both deterministic demo mode and benchmark/stress offline mode. The command-line interface supports a default demo as well as benchmark and stress modes that print human-readable summaries and tables for analysis. This report details the design, implementation, and evaluation of the project and connects the observed behaviors to OS-level concurrency concepts.

\section{Related Works}
Concurrency control has been a central operating-systems concern since early multiprogramming. Seminal work on mutual exclusion and synchronization, including Dijkstra's and Lamport's classic algorithms, established the need for explicit coordination when multiple execution contexts share resources \citep{dijkstra1965,lamport1978}. Practical guidance on programming with threads reinforced these principles for real systems \citep{birrell1989}. Monitors and condition variables, popularized in systems like Mesa, formalized a practical pattern: a shared state protected by a lock, with threads waiting on conditions that are re-checked after waking \citep{hoare1974,lampson1980}. MCRoCS uses this pattern directly, with \texttt{Mutex}-guarded state and \texttt{Condvar} waits for both task queues and zone access \citep{rustmutex,rustcondvar}. This mirrors the approach described in modern OS texts that emphasize correct locking, short critical sections, and explicit wait conditions \citep{arpaci2018,silberschatz2018,tanenbaum2015}.

Task queues are a common concurrency primitive in systems software, appearing in thread pools, work-stealing schedulers, and I/O dispatch loops \citep{arpaci2018,silberschatz2018}. In many high-performance systems, lock-free or multi-queue designs reduce contention, but they are more complex to reason about \citep{herlihy2020}. For the project, the requirement is correctness and clarity rather than maximum throughput. A single FIFO queue guarded by a mutex and condition variable provides strong safety properties: tasks are consumed at most once, and consumers can block efficiently until work arrives \citep{hoare1974}. This is consistent with the ``producer-consumer'' model in classical synchronization literature and is a well-understood pattern for teaching concurrency \citep{downey2016}. Modern practitioner-focused texts likewise emphasize explicit synchronization for shared state in concurrent programs \citep{williams2019}.

Resource coordination often reduces to mutual exclusion on a namespace of shared resources \citep{silberschatz2018,tanenbaum2015}. The zone controller in the project is conceptually similar to a lock table or resource allocator: each zone has at most one owner at a time \citep{gray1993}. This resembles database lock managers and OS-level resource maps where ownership must be unambiguous. In the current implementation, the configured zones are stored in a fixed, direct-indexed slot table, which keeps ownership explicit while avoiding map lookups on the hot path. The use of a condition variable to wait for zone availability aligns with monitor-based coordination patterns \citep{hoare1974,lampson1980}. It is also consistent with the trade-offs discussed in modern concurrency references: coarse-grained locking simplifies correctness reasoning but can limit parallelism if the lock is held too long \citep{herlihy2020}. In the project, each zone acquisition holds the lock only long enough to check and update the ownership slot, minimizing contention.

Health monitoring through heartbeats is common in distributed systems and fault-tolerant designs \citep{lynch1996,chandra1996}. In such systems, nodes periodically emit a signal that a watcher uses to infer liveness. Failure detection is imperfect and depends on timeout selection; shorter timeouts detect failures faster but increase false positives \citep{lynch1996}. The project adopts a simple timeout-based scheme suitable for a single-process simulation. The monitor maintains a last-seen timestamp per robot and marks robots offline once their heartbeat exceeds a configured duration. This is consistent with classic failure detector models and highlights the OS-level idea of watchdog timers \citep{chandra1996}.

Compared with more sophisticated approaches---such as lock-free queues, adaptive timeouts, or priority-aware scheduling---the project intentionally chooses minimal mechanisms. This aligns with the project's pedagogical goals: by using straightforward, well-documented primitives, it becomes easier to explain the safety invariants and to provide testable evidence of correctness. The design therefore emphasizes clarity and correctness over performance, while still enabling quantitative benchmarking.

\section{Implementation}
\subsection{Architecture overview}
MCRoCS is organized around a CLI entry point and three execution modes (demo, benchmark, stress). The core modules implement task queueing, zone access control, and health monitoring. The simulation driver (\texttt{sim.rs}) wires these components, spawns worker threads, and collects metrics. Logging is gated to debug builds, while release mode keeps output concise for measurement and grading.

\begin{figure}[H]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
    box/.style={draw, rounded corners, align=center, inner sep=4pt, font=\small},
    decision/.style={draw, diamond, aspect=2.0, align=center, inner sep=1pt, font=\small},
    arr/.style={-{Latex[length=2mm]}, thick},
    darr/.style={-{Latex[length=2mm]}, thick, dashed}
  ]
    % Entry and routing
    \node[box] (cli) at (0,0) {CLI\\\texttt{main.rs}};
    \node[decision] (mode) at (2.4,0) {Mode};
    \node[box] (demo)   at (5.0,  1.7) {Demo\\\texttt{run\_demo()}};
    \node[box] (bench)  at (5.0,  0.0) {Benchmark\\\texttt{run\_benchmark()}};
    \node[box] (stress) at (5.0, -1.7) {Stress\\\texttt{run\_stress()}};
    \node[box, text width=22mm] (log) at (0, -2.4) {\texttt{log\_dev!}\\(debug only)};

    % Three separate service nodes
    \node[box] (tq) at (9.0,  1.7) {TaskQueue};
    \node[box] (za) at (9.0,  0.0) {ZoneAccess};
    \node[box] (hm) at (9.0, -1.7) {HealthMonitor};

    % Sync-detail nodes (one per service)
    \node[box, text width=40mm] (tqsync) at (13.2,  1.7)
      {\texttt{Mutex\textless VecDeque\textless Task\textgreater\textgreater}\\+ Condvar};
    \node[box, text width=46mm] (zasync) at (13.2,  0.0)
      {\texttt{Mutex\textless Vec\textless Option\textless}\\\texttt{RobotId\textgreater\textgreater\textgreater}\\+ \texttt{Vec\textless Condvar\textgreater}\ (1-indexed per-zone)};
    \node[box, text width=42mm] (hmsync) at (13.2, -1.7)
      {\texttt{Mutex\textless HealthState\textgreater}\\\texttt{heartbeats: HashMap}\\\texttt{offline: HashSet}};

    % Routing arrows
    \draw[arr] (cli)  -- (mode);
    \draw[arr] (mode) -- (demo);
    \draw[arr] (mode) -- (bench);
    \draw[arr] (mode) -- (stress);
    \draw[arr] (stress) -- (bench);

    % Demo → each service
    \draw[arr] (demo) -- (tq);
    \draw[arr] (demo) -- (za);
    \draw[arr] (demo) -- (hm);

    % Bench → each service
    \draw[arr] (bench) -- (tq);
    \draw[arr] (bench) -- (za);
    \draw[arr] (bench) -- (hm);

    % Service → sync detail
    \draw[arr] (tq) -- (tqsync);
    \draw[arr] (za) -- (zasync);
    \draw[arr] (hm) -- (hmsync);

    % Debug-only log coupling (dashed)
    \draw[darr] (demo.west)  to[out=200, in=80]  (log.north);
    \draw[darr] (bench.west) to[out=210, in=10]  (log.east);
  \end{tikzpicture}
  }
  \caption{Architecture overview: CLI routes to execution mode; \texttt{run\_demo()} and \texttt{run\_benchmark()} each use all three core services independently. \texttt{run\_stress()} delegates to \texttt{run\_benchmark()}. Each service has its own synchronization primitive (shown at right). Dashed arrows indicate debug-only coupling (\texttt{log\_dev!} is a no-op in release builds). The zone controller uses direct-indexed per-zone state, and the health monitor keeps heartbeat and offline state under one mutex.}
  \label{fig:architecture}
\end{figure}

\subsection{Task queue}
The task queue uses a \texttt{Mutex<\allowbreak VecDeque<\allowbreak Task>>} to serialize access to an internal FIFO, and a \texttt{Condvar} to block consumers when the queue is empty \citep{rustmutex,rustcondvar}. This design ensures that each task is consumed at most once, and it prevents busy waiting. The critical section is minimal: a consumer holds the lock only long enough to check for a task and pop it. When the queue is empty, the thread waits on the condition variable, which releases the lock and re-acquires it on wake.

\noindent\textbf{Critical code path (blocking pop):}
\begin{lstlisting}
pub fn pop_blocking_or_closed(&self) -> Option<Task> {
    let mut guard = self.inner.lock().expect("task queue mutex poisoned");
    loop {
        if let Some(task) = guard.queue.pop_front() {
            return Some(task);
        }
        if guard.closed {
            return None;
        }
        guard = self.available.wait(guard).expect("condvar wait failed");
    }
}
\end{lstlisting}

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    node distance=7mm and 10mm,
    box/.style={draw, rounded corners, align=center, inner sep=4pt, font=\small},
    decision/.style={draw, diamond, aspect=2.2, align=center, inner sep=1pt, font=\small},
    arr/.style={-{Latex[length=2mm]}, thick}
  ]
    %% ---- Left column: Producer (push) ----
    \node[box] (p0) at (-3.2, 0)    {\textbf{Producer}\\\texttt{push(task)}};
    \node[box] (p1) at (-3.2,-1.2)  {Lock Mutex};
    \node[decision] (p2) at (-3.2,-2.6) {closed?};
    \node[box] (p3) at (-5.2,-4.0)  {Return\\Err(task)};
    \node[box] (p4) at (-1.8,-4.0)  {push\_back\\to VecDeque};
    \node[box] (p5) at (-1.8,-5.3)  {notify\_one};
    \node[box] (p6) at (-1.8,-6.6)  {Unlock};

    \draw[arr] (p0) -- (p1);
    \draw[arr] (p1) -- (p2);
    \draw[arr] (p2) -- node[above left, font=\small]{yes} (p3);
    \draw[arr] (p2) -- node[above right, font=\small]{no}  (p4);
    \draw[arr] (p4) -- (p5);
    \draw[arr] (p5) -- (p6);

    %% ---- Right column: Consumer (pop_blocking_or_closed) ----
    \node[box] (c0) at (3.5, 0)    {\textbf{Consumer}\\\texttt{pop\_blocking\_or\_closed()}};
    \node[box] (c1) at (3.5,-1.2)  {Lock Mutex};
    \node[decision] (c2) at (3.5,-2.6) {task\\available?};
    \node[box] (c3) at (1.4,-4.0)  {pop\_front\\return Some(task)};
    \node[decision] (c4) at (5.0,-4.0) {closed?};
    \node[box] (c5) at (3.8,-5.3)  {return None};
    \node[box] (c6) at (6.4,-5.3)  {Condvar::wait\\(releases lock)};

    \draw[arr] (c0) -- (c1);
    \draw[arr] (c1) -- (c2);
    \draw[arr] (c2) -- node[above left, font=\small]{yes} (c3);
    \draw[arr] (c2) -- node[above right, font=\small]{no}  (c4);
    \draw[arr] (c4) -- node[above left, font=\small]{yes}  (c5);
    \draw[arr] (c4) -- node[above right, font=\small]{no}  (c6);
    % wait loops back to the task-available check
    \draw[arr] (c6.east) -- ++(0.5,0) |- (c2.east);

    %% ---- Vertical separator ----
    \draw[dashed, gray] (0, 0.6) -- (0, -7.0);
  \end{tikzpicture}
  \caption{TaskQueue producer and consumer paths: \texttt{push} notifies one consumer; blocking \texttt{pop\_blocking\_or\_closed} waits until a task arrives or the queue closes.}
  \label{fig:tq-flow}
\end{figure}

The queue also supports \texttt{try\_pop} for non-blocking access and a \texttt{close} operation that wakes all blocked consumers. Tests validate that tasks are consumed exactly once across multiple threads, that blocking consumers wake on push, and that closure unblocks waiting threads. These tests directly target the invariants required by the project: single assignment of tasks and safe concurrent access.

\subsection{Zone access control}
The zone controller enforces exclusive occupancy per zone using a \texttt{Mutex<Vec<Option<RobotId>>>} and a vector of per-zone \texttt{Condvar}s \citep{rustmutex,rustcondvar}. Zones are 1-indexed and validated against the configured zone count, so each zone maps directly to one ownership slot and one condvar. This keeps the representation simple and removes hash-map lookup overhead from the hot path while preserving the same mutual-exclusion behavior. When a zone becomes available, the owner calls \texttt{notify\_one} on that zone's condvar, waking exactly one contender rather than all waiters.

\noindent\textbf{Critical code path (acquire):}
\begin{lstlisting}
pub fn acquire(&self, zone: ZoneId, robot: RobotId) {
    let idx = self.zone_index(zone);
    let mut guard = self.occupied.lock().expect("zone mutex poisoned");
    loop {
        if guard[idx].is_none() {
            guard[idx] = Some(robot);
            return;
        }
        guard = self.condvars[idx].wait(guard).expect("condvar wait failed");
    }
}
\end{lstlisting}

This approach keeps the lock scope short and avoids deadlocks because only one lock is held at a time. Per-zone condvars eliminate unnecessary wakeups when zones are independent, reducing contention under high concurrency. The design is intentionally conservative: it prioritizes correctness and explicit ownership tracking. Unit tests spawn multiple contenders for a single zone and verify that the observed occupancy never exceeds one; additional tests check that the occupied-zone snapshot matches current ownership and that out-of-range zones are rejected in debug builds. In release builds, non-owner release attempts are logged and rejected, preserving safety even without debug assertions.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    node distance=9mm and 12mm,
    box/.style={draw, rounded corners, align=center, inner sep=4pt, font=\small},
    decision/.style={draw, diamond, aspect=2.2, align=center, inner sep=1pt, font=\small},
    arr/.style={-{Latex[length=2mm]}, thick}
  ]
    \node[box] (a) {Robot requests zone};
    \node[decision, below=of a] (b) {Zone\\occupied?};
    \node[box, below left=of b] (c) {Insert owner\\and enter zone};
    \node[box, below right=of b] (d) {Wait on\\condvar};
    \node[box, below=of c] (e) {Robot performs work};
    \node[box, below=of e] (f) {Release by owner};
    \node[box, below=of f] (g) {Remove owner\\entry};
    \node[box, below=of g] (h) {notify\_one\\--- wake one waiter};

    \draw[arr] (a) -- (b);
    \draw[arr] (b) -- node[above left]{no} (c);
    \draw[arr] (b) -- node[above right]{yes} (d);
    \draw[arr] (d) -- ++(0,-7mm) -| (b.east);
    \draw[arr] (c) -- (e);
    \draw[arr] (e) -- (f);
    \draw[arr] (f) -- (g);
    \draw[arr] (g) -- (h);
  \end{tikzpicture}
  \caption{Zone access control logic (TikZ adaptation of the implementation flow).}
  \label{fig:zone-flow}
\end{figure}

\subsection{Health monitor}
The health monitor uses a single-lock design: a \texttt{Mutex<HealthState>} that contains both a \texttt{HashMap<RobotId, Instant>} for heartbeat timestamps and a \texttt{HashSet<RobotId>} for the offline set \citep{rustmutex}. This keeps heartbeat updates and offline detection inside one coherent critical section and removes the need for cross-lock coordination. The implementation relies on monotonic timestamps for timeout comparison \citep{rustinstant}.

\noindent\textbf{Critical code path (offline detection):}
\begin{lstlisting}
pub fn detect_offline(&self, timeout: Duration) -> HashSet<RobotId> {
    let now = Instant::now();
    let mut guard = self.inner.lock().expect("health monitor mutex poisoned");
    let overdue: Vec<_> = guard
        .heartbeats
        .iter()
        .filter_map(|(&robot, &last)| {
            if now.duration_since(last) > timeout { Some(robot) } else { None }
        })
        .collect();
    for robot in overdue {
        guard.offline.insert(robot);
    }
    guard.offline.clone()
}
\end{lstlisting}

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
    state/.style={draw, circle, minimum size=22mm, align=center, font=\small},
    arr/.style={-{Latex[length=2mm]}, thick}
  ]
    % States
    \node[state] (unreg)   at (0,   0)  {Unregistered};
    \node[state] (online)  at (5.0, 0)  {Online};
    \node[state] (offline) at (10.0, 0) {Offline};

    % Initial pseudo-state
    \node[draw, fill, circle, minimum size=3mm, inner sep=0pt] (init) at (-2.2, 0) {};
    \draw[arr] (init) -- (unreg);

    % Unregistered → Online
    \draw[arr] (unreg) -- node[above, font=\small]
      {\texttt{register\_robot()} or \texttt{heartbeat()}} (online);

    % Online self-loop (top)
    \draw[arr] (online) to[out=60, in=120, looseness=5]
      node[above, font=\small] {\texttt{heartbeat()} — resets timestamp} (online);

    % Online → Offline
    \draw[arr] (online) to[out=-30, in=210]
      node[above, font=\small] {\texttt{detect\_offline()} — heartbeat expired}
      (offline);

    % Offline → Online
    \draw[arr] (offline) to[out=150, in=30]
      node[below, font=\small] {\texttt{heartbeat()} — clears offline mark}
      (online);
  \end{tikzpicture}
  \caption{HealthMonitor robot state transitions. A robot enters \textit{Online} on first heartbeat, transitions to \textit{Offline} when \texttt{detect\_offline()} finds its timestamp overdue, and recovers to \textit{Online} on the next heartbeat.}
  \label{fig:hm-states}
\end{figure}

Tests ensure that robots are marked offline after timeouts and that a new heartbeat clears the offline status. A test-only hook allows deterministic verification without sleeping, preventing flaky test behavior.

\subsection{Simulation and CLI}

\begin{figure}[H]
  \centering
  \resizebox{0.97\linewidth}{!}{%
  \begin{tikzpicture}[
    arr/.style={-{Latex[length=2mm]}, thick},
    note/.style={draw, rounded corners, align=center, inner sep=3pt, font=\small, fill=white},
    tl/.style={dashed, gray}
  ]
    % Participant headers
    \node[font=\small\bfseries, align=center] at (0,    0.7) {Robot 0\\Robot 2};
    \node[font=\small\bfseries, align=center] at (2.6,  0.7) {Robot 1\\(offline target)};
    \node[font=\small\bfseries, align=center] at (5.6,  0.7) {TaskQueue /\\ZoneAccess};
    \node[font=\small\bfseries, align=center] at (8.4,  0.7) {Health\\Monitor};
    \node[font=\small\bfseries, align=center] at (11.0, 0.7) {Main\\Thread};

    % Timelines
    \draw[tl] (0,    0) -- (0,    -11.2);
    \draw[tl] (2.6,  0) -- (2.6,  -11.2);
    \draw[tl] (5.6,  0) -- (5.6,  -11.2);
    \draw[tl] (8.4,  0) -- (8.4,  -11.2);
    \draw[tl] (11.0, 0) -- (11.0, -11.2);

    % par block bracket
    \draw[dashed, thin] (-0.6, -0.5) rectangle (6.2, -2.2);
    \node[font=\scriptsize\itshape, anchor=north west] at (-0.55, -0.55) {par};

    % 1. Concurrent pop_blocking_or_closed
    \draw[arr] (0,   -1.1) -- node[above, font=\scriptsize] {\texttt{pop\_blocking\_or\_closed()}} (5.6, -1.1);
    \draw[arr] (2.6, -1.7) -- node[below, font=\scriptsize] {\texttt{pop\_blocking\_or\_closed()}} (5.6, -1.7);

    % 2. acquire(zone)
    \draw[arr] (0,   -3.0) -- node[above, font=\scriptsize] {\texttt{acquire(zone)} \textit{(blocks if busy)}} (5.6, -3.0);
    \draw[arr] (2.6, -3.5) -- node[below, font=\scriptsize] {\texttt{acquire(zone)}} (5.6, -3.5);

    % 3. R0/R2 heartbeat after each task
    \draw[arr] (0,   -4.8) -- node[above, font=\scriptsize] {\texttt{heartbeat()} after each task} (8.4, -4.8);

    % 4. R1 heartbeat — tasks 1 and 2 only
    \draw[arr] (2.6, -5.5) -- node[above, font=\scriptsize] {\texttt{heartbeat()} after tasks 1 and 2 only} (8.4, -5.5);

    % 5. R1 stops heartbeats note
    \node[note] at (2.6, -6.4) {\scriptsize stops heartbeats\\[-1pt]\scriptsize after task 2};

    % 6. detect_offline self-call on HM (right-side loop)
    \draw[arr] (8.4, -7.1) to[out=-30, in=30, looseness=7]
      node[right, font=\scriptsize, align=left] {\texttt{detect\_offline(200ms)}\\every 50ms} (8.4, -7.7);

    % Divider: robot threads join
    \draw[dotted, thick] (-0.4, -8.5) -- (11.4, -8.5);
    \node[font=\scriptsize\itshape] at (5.5, -8.3) {robot threads join};

    % 7. Main keepalive heartbeats for R0, R2
    \draw[arr] (11.0, -9.2) -- node[above, font=\scriptsize] {\texttt{heartbeat()} keepalive for R0, R2} (8.4, -9.2);

    % 8. wait_for_specific_offline
    \draw[arr] (11.0, -10.2) -- node[above, font=\scriptsize]
      {\texttt{wait\_for\_specific\_offline}(target=1, 200ms timeout)} (8.4, -10.2);

    % 9. HM → Main: offline detected
    \draw[arr] (8.4, -11.0) -- node[above, font=\scriptsize] {robot 1 offline detected} (11.0, -11.0);

  \end{tikzpicture}
  }
  \caption{Demo sequence: robots concurrently fetch tasks and acquire zones; Robot 1 stops heartbeats after task 2, triggering offline detection.}
  \label{fig:robot-sequence}
\end{figure}

The CLI (\texttt{main.rs}) provides three entry points: demo, benchmark, and stress. In demo mode, robot 1 is the deterministic offline target: it stops heartbeats mid-run while non-target robots are kept alive during final timeout verification. The summary reports \texttt{offline\_target}, \texttt{detected}, per-robot task counts, and zone-safety fields. Shared ownership of core services uses reference-counted pointers \citep{rustarc}.

Benchmarks and stress runs reuse the same core logic and report elapsed time, throughput, average zone wait, CPU usage (when available), and validation flags (duplicate tasks, zone violations). In offline benchmark/stress mode, the system waits for any offline detection after workload completion, so the final offline count may exceed one depending on timing. In normal benchmark/stress mode, the background health-monitor thread is not spawned, which avoids paying offline-detection overhead when liveness is not part of the measured scenario. These runs are tuned for repeatable, observable output rather than realism.

From an implementation perspective, the CLI also enforces strict input validation to keep experiments reproducible. Invalid values (for example zero robots or zero zones) are rejected with a clear usage message, and stress lists can preserve defaults through explicit placeholders. This reduces ambiguity during evaluation because graders can run the same commands and receive predictable behavior and diagnostics. Together with deterministic demo fields and the formatted benchmark/stress summaries shown by the current CLI, these interface decisions improve traceability between source code, runtime evidence, and report claims.

\subsection{Synchronization strategy and invariants}
Each component uses a small, fixed set of synchronization primitives with a consistent lock order, avoiding lock cycles and eliminating deadlock risks. The task queue and zone controller each center around a single mutex plus condition variables to coordinate waiting without busy loops. The zone controller uses direct-indexed per-zone condvars so that robots waiting on independent zones do not wake each other, reducing spurious contention. The health monitor uses a single \texttt{Mutex<HealthState>} so heartbeats and offline detection update one coherent structure without cross-lock races. Lock scopes are kept tight to reduce contention. The benchmark harness tracks duplicate tasks via a fixed seen-table keyed by task ID, which avoids channel traffic and keeps validation overhead predictable. Atomics are used only for lightweight counters in the simulation and do not replace the fundamental locks that protect shared structures.

The design enforces the required safety invariants:
\begin{itemize}[leftmargin=*]
  \item Each task is removed from the queue at most once.
  \item A zone is owned by at most one robot at a time.
  \item Robots are marked offline if their heartbeat exceeds a timeout.
  \item Critical sections are short, and lock ordering remains consistent to avoid deadlock.
\end{itemize}

\section{Benchmark}
\subsection{Methodology}
Benchmarks were measured on 2026-03-20 on macOS (Darwin 25.3.0, arm64) using \texttt{rustc 1.93.0}. Before collecting performance data, the required correctness gates were re-run: \texttt{cargo build --release} completed successfully and \texttt{cargo test} passed all 17 tests. All published numbers were then collected from the release binary (\texttt{./target/release/project\_blaze bench ...}) with the \texttt{validate} flag enabled, which keeps duplicate-task and zone-violation checks active during measurement.

To reduce coupling between rows, each published configuration was executed as five isolated benchmark runs rather than copied from one long stress sweep. The tables report mean \(\pm\) sample standard deviation for elapsed time, throughput, average zone wait, and CPU time. CPU user/system values come from a Unix \texttt{getrusage} delta and therefore support relative comparison on this machine rather than cross-platform claims. The benchmark output also reports \texttt{max\_occupancy}, \texttt{zone\_violation}, \texttt{duplicate\_tasks}, and \texttt{offline\_robots}, so each timing result is paired with an explicit correctness signal. Offline-mode data is interpreted as ``at least one robot detected offline''; unlike the demo, the benchmark does not target a fixed robot identity.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=9mm and 12mm,
  box/.style={draw, rounded corners, align=center, inner sep=4pt, font=\small},
  decision/.style={draw, diamond, aspect=2.2, align=center, inner sep=1pt, font=\small},
  arr/.style={-{Latex[length=2mm]}, thick}
]
  \node[box] (a) {Start benchmark/stress run};
  \node[decision, below=of a] (b) {offline-demo\\enabled?};
  \node[box, below left=of b] (c) {Normal run\\all robots keep heartbeats};
  \node[box, below right=of b] (d) {Spawn monitor thread\\robot 0 stops heartbeats\\after \texttt{tasks\_per\_robot/2}};
  \node[box, below=11mm of b] (e) {Workers finish tasks};
  \node[box, below=of e] (f) {\texttt{wait\_for\_offline(any,}\\\texttt{timeout=500ms, max\_wait=1000ms)}};
  \node[box, below=of f] (g) {Collect\\\texttt{offline\_robots}};
  \node[box, below=of g] (h) {Emit benchmark/stress row};

  \draw[arr] (a) -- (b);
  \draw[arr] (b) -- node[above left]{no} (c);
  \draw[arr] (b) -- node[above right]{yes} (d);
  \draw[arr] (c) |- (e);
  \draw[arr] (d) |- (e);
  \draw[arr] (e) -- (f);
  \draw[arr] (f) -- (g);
  \draw[arr] (g) -- (h);
\end{tikzpicture}
\caption{Offline benchmark semantics, updated to match the current implementation and \texttt{DIAGRAMS.md}.}
\label{fig:bench-offline-flow}
\end{figure}

\subsection{Results}
Table~\ref{tab:bench-samples} summarizes four representative scenarios. The baseline and high-contention cases use the same work per task, so their nearly identical throughput is expected: with only two zones, the extra robots mainly increase waiting rather than useful parallel work. Expanding from two zones to four almost doubles throughput and keeps average wait far below the 8-robot/2-zone case, showing that the synchronization design allows additional concurrency when independent resources exist. Offline mode has the lowest throughput because the run intentionally waits for liveness timeout detection after the workload completes; this cost is a property of the measurement scenario, not a safety failure.

\begin{table}[H]
\centering
\small
\setlength{\tabcolsep}{5pt}
\resizebox{\linewidth}{!}{%
\begin{tabular}{lrrrr}
\toprule
Scenario & Elapsed (ms) & Throughput (tasks/s) & Avg wait (\(\mu\)s) & CPU sys (s) \\
\midrule
4 robots, 25 tasks, 2 zones & 312.20 \(\pm\) 4.09 & 320.35 \(\pm\) 4.15 & 4254.90 \(\pm\) 1234.38 & 0.0016 \(\pm\) 0.0001 \\
8 robots, 25 tasks, 2 zones & 624.60 \(\pm\) 10.81 & 320.28 \(\pm\) 5.47 & 12143.30 \(\pm\) 1236.06 & 0.0033 \(\pm\) 0.0001 \\
8 robots, 25 tasks, 4 zones & 337.20 \(\pm\) 9.86 & 593.53 \(\pm\) 17.66 & 4951.34 \(\pm\) 288.69 & 0.0038 \(\pm\) 0.0001 \\
4 robots, 25 tasks, 2 zones, offline & 707.60 \(\pm\) 48.53 & 141.91 \(\pm\) 10.71 & 4489.61 \(\pm\) 1010.36 & 0.0020 \(\pm\) 0.0002 \\
\bottomrule
\end{tabular}
}
\caption{Five-run release-mode benchmark summary. In all rows, \texttt{zone\_violation=false} and \texttt{duplicate\_tasks=false}.}
\label{tab:bench-samples}
\end{table}

\begin{table}[H]
\centering
\small
\setlength{\tabcolsep}{5pt}
\begin{tabular}{llrrr}
\toprule
Study & Config & Elapsed (ms) & Throughput (tasks/s) & Avg wait (\(\mu\)s) \\
\midrule
Robot/zone & 1 robot, 1 zone & 62.00 \(\pm\) 0.00 & 161.29 \(\pm\) 0.00 & 0.00 \(\pm\) 0.00 \\
Robot/zone & 1 robot, 2 zones & 61.40 \(\pm\) 0.89 & 162.89 \(\pm\) 2.40 & 0.00 \(\pm\) 0.00 \\
Robot/zone & 2 robots, 1 zone & 123.40 \(\pm\) 1.67 & 162.10 \(\pm\) 2.21 & 3082.60 \(\pm\) 71.94 \\
Robot/zone & 2 robots, 2 zones & 61.60 \(\pm\) 0.55 & 324.70 \(\pm\) 2.90 & 18.15 \(\pm\) 18.60 \\
Robot/zone & 4 robots, 1 zone & 245.40 \(\pm\) 1.82 & 163.00 \(\pm\) 1.21 & 9284.91 \(\pm\) 252.36 \\
Robot/zone & 4 robots, 2 zones & 123.80 \(\pm\) 0.84 & 323.11 \(\pm\) 2.18 & 3886.03 \(\pm\) 753.62 \\
Task load & 4 robots, 10 tasks, 2 zones & 124.80 \(\pm\) 3.11 & 320.67 \(\pm\) 7.83 & 4951.00 \(\pm\) 1120.75 \\
Task load & 4 robots, 25 tasks, 2 zones & 311.40 \(\pm\) 1.67 & 321.14 \(\pm\) 1.72 & 4184.38 \(\pm\) 1093.14 \\
Task load & 4 robots, 50 tasks, 2 zones & 632.40 \(\pm\) 15.52 & 316.40 \(\pm\) 7.59 & 5311.44 \(\pm\) 720.11 \\
\bottomrule
\end{tabular}
\caption{Five-run scalability and task-load comparisons. \texttt{max\_occupancy} always matched the zone count, and no row produced violations or duplicates.}
\label{tab:scalability}
\end{table}

\begin{figure}[H]
\centering
\begin{subfigure}[t]{0.49\linewidth}
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\linewidth,
  height=5.6cm,
  ybar,
  bar width=10pt,
  ymin=0,
  ylabel={Throughput (tasks/s)},
  symbolic x coords={4R/2Z,8R/2Z,8R/4Z,4R/2Z offline},
  xtick=data,
  x tick label style={rotate=20,anchor=east,font=\scriptsize},
  nodes near coords,
  nodes near coords align={vertical},
  every node near coord/.append style={font=\scriptsize},
]
\addplot[fill=cyan!55] coordinates {
  (4R/2Z,320.35)
  (8R/2Z,320.28)
  (8R/4Z,593.53)
  (4R/2Z offline,141.91)
};
\end{axis}
\end{tikzpicture}
\caption{Headline scenarios.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.49\linewidth}
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\linewidth,
  height=5.6cm,
  ymin=0,
  xlabel={Robots},
  ylabel={Throughput (tasks/s)},
  xtick={1,2,4},
  legend style={at={(0.5,-0.25)},anchor=north,legend columns=2,font=\scriptsize},
  grid=major,
]
\addplot+[mark=*] coordinates {(1,161.29) (2,162.10) (4,163.00)};
\addplot+[mark=square*] coordinates {(1,162.89) (2,324.70) (4,323.11)};
\legend{1 zone, 2 zones}
\end{axis}
\end{tikzpicture}
\caption{Robot/zone scaling.}
\end{subfigure}
\caption{LaTeX-native charts replacing the earlier ASCII visualization.}
\label{fig:bench-charts}
\end{figure}

\subsection{Interpretation}
The fresh measurements support three conclusions. First, correctness remained stable throughout the evaluation: duplicate-task detection never triggered, zone exclusivity never failed, and offline mode consistently reported one offline robot in the designated liveness scenario. Second, contention is the dominant performance bottleneck. With one zone, throughput stays near 162 tasks/s as robots increase because threads spend more time waiting for the same protected resource. With two zones, throughput roughly doubles at 2 and 4 robots, which confirms that the synchronization design is not serializing unrelated zone accesses. Third, task-load growth is close to linear in elapsed time while throughput stays near 320 tasks/s for the 4-robot/2-zone configuration; this suggests that the coordination costs are largely steady-state rather than startup-driven. These trends are exactly what a minimal mutex-and-condition-variable design should exhibit: safe sharing first, then parallel speedup only when independent zones are available.

The CPU-time columns reinforce that interpretation. System time rises from about 0.0016\,s in the baseline case to 0.0038\,s in the 8-robot/4-zone case, which is consistent with more successful lock acquisitions and zone transitions rather than busy waiting. Offline mode is different: its average zone wait stays in the same order of magnitude as the normal 4-robot/2-zone run, but total elapsed time expands because the harness deliberately waits for timeout confirmation after the work has completed. A cross-check with the CLI stress sweep showed the same qualitative pattern as the repeated single-run measurements, so the reported tables capture stable behavior rather than a one-off schedule artifact.

\section{Discussion}
The core goal of the project is correctness in concurrency, and the implementation meets this requirement by applying straightforward synchronization patterns. The task queue demonstrates a classic producer-consumer solution with a mutex and condition variable, yielding deterministic, single-consumer behavior. The zone controller models exclusive resource ownership, ensuring that a robot must explicitly acquire and release a zone, a pattern familiar from lock managers in systems and databases. The health monitor illustrates a watchdog-style approach to liveness detection and shows how timeout-driven logic can be implemented safely with a single mutex-protected state for heartbeat timestamps and offline markers.

The results highlight predictable trade-offs. When the number of robots increases without a matching increase in zones, throughput improvements are limited by contention and by the time each robot holds a zone. This is visible in the higher average zone wait times for configurations with many robots and few zones. Conversely, adding zones improves throughput and reduces wait time, demonstrating that the synchronization scheme allows parallelism when resources are available. CPU usage remains low in these microbenchmarks because most of the wall time is spent sleeping to simulate work, which is appropriate for a coordination-focused simulation.

The benchmark instrumentation confirms that per-zone occupancy never exceeded one in the sampled runs, which is consistent with the \texttt{ZoneAccess} unit tests and with the single-lock acquisition logic. This provides direct empirical evidence that the synchronization design preserves mutual exclusion under contention.

Another trade-off involves the heartbeat timeout. A shorter timeout improves detection speed but risks marking a slow robot offline. The demo uses short timeouts to make the offline event visible, whereas a real deployment would choose a longer threshold and possibly incorporate adaptive timing. The current design keeps this parameter explicit and configurable at the simulation level, which makes the timing assumptions transparent and easier to justify in a report.

From a software engineering perspective, Rust's ownership model and standard synchronization primitives promote safe concurrency. The use of \texttt{Arc} for shared ownership together with \texttt{Mutex} and \texttt{Condvar} avoids data races without requiring unsafe code. The design avoids inconsistent lock ordering, which eliminates deadlock cycles and keeps critical sections short. Recent internal optimizations preserved this simplicity: direct-indexed zone ownership, a single-lock health monitor, and a lower-overhead validation path improved benchmark behavior without introducing feature creep. This aligns with the project's decision rule of correctness over performance and supports the learning outcome that synchronization should be explicit and simple when possible.

Alternative primitives were considered but were less appropriate for this workload. An \texttt{RwLock} would add mode-switching complexity without much benefit because queue pops, zone ownership changes, and heartbeat updates are all write-heavy operations. Message-passing channels could replace the task queue, but they would not remove the need for shared zone ownership and timeout bookkeeping. For this project, a uniform \texttt{Mutex}+\texttt{Condvar} approach kept the coordination story easier to verify, benchmark, and explain \citep{rustmutex,rustcondvar,herlihy2020}.

A limitation of the current evaluation is that benchmarks are short and focused on synchronization behavior rather than sustained load. Longer runs would reduce noise in throughput and CPU measurements, and repeated trials would allow confidence intervals. Additionally, the simulation uses a fixed work duration and deterministic task-to-zone mapping, which simplifies analysis but does not capture heterogeneous task costs or dynamic zone selection. These constraints are acceptable for the minimal scope but should be revisited if the system evolves toward more realistic scheduling. From a dependability perspective, the current evaluation focuses on safety and correctness rather than on formal reliability or availability metrics \citep{avizienis2004}.

Overall, the project demonstrates that minimal, well-structured synchronization can produce a correct and observable coordination layer. The system's behavior is easy to reason about, and the design choices map cleanly to OS concurrency principles. The remaining gaps are primarily in the rigor of benchmark instrumentation and in the breadth of performance evaluation, rather than in the correctness of the core synchronization mechanisms.

\section{Conclusion and Future Work}
MCRoCS achieves its goal of implementing a lightweight OS core that safely coordinates multiple robot threads. It delivers the required task queue, zone access control, and health monitor with clear synchronization boundaries and minimal shared state. The demo shows multiple robots fetching tasks concurrently, exclusive zone occupancy, and an offline robot detected through heartbeat timeouts. Unit tests and integration tests provide direct evidence of correctness for each required component, and benchmark runs quantify throughput and waiting behavior under contention.

The project's minimal design was a deliberate choice to prioritize correctness and clarity. Using mutexes and condition variables makes the concurrency model explicit and easy to analyze, while avoiding deadlock and complex scheduling policies. The architecture is modular, with each component encapsulating its own synchronization, which simplifies reasoning and supports extension. These properties align with the learning outcomes for concurrency control, synchronization, and coordination.

Future work can build on this foundation in several directions. First, longer and repeated performance runs should be added to reduce measurement noise and to generate statistically meaningful metrics and graphs. Second, the simulation could incorporate heterogeneous task durations, dynamic zone selection, and injected delays to model more realistic operational conditions. Third, the health monitor could be extended with adaptive timeout policies or a suspicion window to balance detection speed with false-positive avoidance. Finally, if the project scope expands, it would be natural to explore advanced scheduling strategies or more scalable queue designs, while preserving the correctness guarantees that the minimal system already demonstrates.

To improve evaluation quality further, benchmark and stress modes could adopt deterministic liveness attribution similar to the demo rather than relying only on ``any offline'' outcomes. In practical terms, a fixed target robot could be selected to stop heartbeats while non-target robots continue sending periodic heartbeats until timeout verification finishes. That extension would make workload-timeout results easier to interpret, because the summary could distinguish the intended offline robot from incidental end-of-run silence.

In its current form, the system is already suitable for the required demo and report deliverables: it has a clear architecture, observable concurrency behavior, and a clean test suite that verifies the three mandatory components. These qualities make it a strong baseline for any future extensions.

\begin{thebibliography}{99}
\bibitem[Arpaci-Dusseau \& Arpaci-Dusseau(2023)]{arpaci2018} Arpaci-Dusseau, R. H., \& Arpaci-Dusseau, A. C. (2023). \textit{Operating Systems: Three Easy Pieces} (Version 1.10). Arpaci-Dusseau Books. Retrieved February 2, 2026, from https://pages.cs.wisc.edu/~remzi/OSTEP/
\bibitem[Birrell(1989)]{birrell1989} Birrell, A. D. (1989). \textit{An Introduction to Programming with Threads}. Digital Equipment Corporation.
\bibitem[Dijkstra(1965)]{dijkstra1965} Dijkstra, E. W. (1965). \textit{Cooperating sequential processes} (EWD 123). E. W. Dijkstra Archive, The University of Texas at Austin. Retrieved March 20, 2026, from https://www.cs.utexas.edu/~EWD/transcriptions/EWD01xx/EWD123.html
\bibitem[Downey(2016)]{downey2016} Downey, A. B. (2016). \textit{The Little Book of Semaphores} (2nd ed.). Green Tea Press.
\bibitem[Hoare(1974)]{hoare1974} Hoare, C. A. R. (1974). Monitors: An operating system structuring concept. \textit{Communications of the ACM, 17}(10), 549--557.
\bibitem[Lampson \& Redell(1980)]{lampson1980} Lampson, B. W., \& Redell, D. D. (1980). Experience with processes and monitors in Mesa. \textit{Communications of the ACM, 23}(2), 105--117.
\bibitem[Lamport(1978)]{lamport1978} Lamport, L. (1978). Time, clocks, and the ordering of events in a distributed system. \textit{Communications of the ACM, 21}(7), 558--565.
\bibitem[Herlihy et~al.(2020)]{herlihy2020} Herlihy, M., Shavit, N., Luchangco, V., \& Spear, M. (2020). \textit{The Art of Multiprocessor Programming} (2nd ed.). Morgan Kaufmann.
\bibitem[Silberschatz et~al.(2018)]{silberschatz2018} Silberschatz, A., Galvin, P. B., \& Gagne, G. (2018). \textit{Operating System Concepts} (10th ed.). Wiley.
\bibitem[Tanenbaum \& Bos(2015)]{tanenbaum2015} Tanenbaum, A. S., \& Bos, H. (2015). \textit{Modern Operating Systems} (4th ed.). Pearson.
\bibitem[Williams(2019)]{williams2019} Williams, A. (2019). \textit{C++ Concurrency in Action} (2nd ed.). Manning.
\bibitem[Klabnik \& Nichols(2022)]{klabnik2022} Klabnik, S., \& Nichols, C. (2022). \textit{The Rust Programming Language} (2nd ed.). No Starch Press.
\bibitem[Rust Project(n.d.-a)]{rustmutex} Rust Project. (n.d.). \textit{std::sync::Mutex}. Retrieved February 2, 2026, from https://doc.rust-lang.org/std/sync/struct.Mutex.html
\bibitem[Rust Project(n.d.-b)]{rustcondvar} Rust Project. (n.d.). \textit{std::sync::Condvar}. Retrieved February 2, 2026, from https://doc.rust-lang.org/std/sync/struct.Condvar.html
\bibitem[Rust Project(n.d.-c)]{rustarc} Rust Project. (n.d.). \textit{std::sync::Arc}. Retrieved February 2, 2026, from https://doc.rust-lang.org/std/sync/struct.Arc.html
\bibitem[Rust Project(n.d.-e)]{rustinstant} Rust Project. (n.d.). \textit{std::time::Instant}. Retrieved February 2, 2026, from https://doc.rust-lang.org/std/time/struct.Instant.html
\bibitem[Gray \& Reuter(1993)]{gray1993} Gray, J., \& Reuter, A. (1993). \textit{Transaction Processing: Concepts and Techniques}. Morgan Kaufmann.
\bibitem[Lynch(1996)]{lynch1996} Lynch, N. (1996). \textit{Distributed Algorithms}. Morgan Kaufmann.
\bibitem[Chandra \& Toueg(1996)]{chandra1996} Chandra, T. D., \& Toueg, S. (1996). Unreliable failure detectors for reliable distributed systems. \textit{Journal of the ACM, 43}(2), 225--267.
\bibitem[Avizienis et~al.(2004)]{avizienis2004} Avizienis, A., Laprie, J.-C., Randell, B., \& Landwehr, C. E. (2004). Basic concepts and taxonomy of dependable and secure computing. \textit{IEEE Transactions on Dependable and Secure Computing, 1}(1), 11--33. https://doi.org/10.1109/TDSC.2004.2
\end{thebibliography}

\end{document}
